A Decision Procedure for String Constraints with String-Integer Conversion and Flat Regular Constraints

### 2021-05-14 FMCAD2021/1_intro.tex
sec6 half done

One more recent approach is flattening~\cite{}.
The idea is to restrict the solution space of string variables to some pattern defined by \emph{flat automata} (see Section~\ref{PA} for more details). 
The major benefit of considering this class is two folds.
First, even under the restriction, often the potential solution space is still infinite, which gives us a higher potential of finding solutions.
For instance, we can find a solution for $x.y \neq z  \wedge |x| > 2000$ under a very simple restriction: all variables are in $a^*$, which $a$ is a character in our alphabet.
Second, and more importantly, because we can convert the membership problem of a flat automaton to a satisfiability problem of Presburgh formula, the class of word equation + regular constraints + length constraints is decidable.

The paper of Abdulla et al. has considered adding string/number conversion constraints to the above class.
The paper proposes a decision procedure with a restricted form of flat automata.
For string/number conversion, their approach projects the solution to string/number conversion to, in terms of numerical solutions, a finite solution space.
In this paper, we propose a complete solution to this problem.
We propose a decision procedure for the class of word equation + regular constraints + length constraints + string/number conversion under arbitrary flat automata.

>>>>

One more recent approach is flattening~\cite{A}.
The idea is to restrict the solution space of string variables to some pattern defined by \emph{flat automata} (see Section~\ref{PA} for more details). 
The major benefit of considering this class is two folds.
First, even under the restriction, often the potential solution space is still infinite, which gives us a higher potential of finding solutions.
For instance, we can find a solution for $x.y \neq z  \wedge |x| > 2000$ under a very simple restriction: all variables are in $a^*$, which $a$ is a character in our alphabet.
Second, and more importantly, because we can convert the membership problem of a flat automaton to a satisfiability problem of Presburgher formula, the class of word equation + regular constraints + length constraints is decidable.

The paper of Abdulla et al. has considered adding string/number conversion constraints to the above class.
The paper proposes a decision procedure with a restricted form of flat automata.
For string/number conversion, their approach projects the solution to string/number conversion to, in terms of numerical solutions, a finite solution space.

In this paper, we propose a complete solution to this problem.
We propose a decision procedure for the class of word equation + regular constraints + length constraints + string/number conversion under arbitrary flat automata (denoted by $\strint$).
The basic idea of our approach can be sketched as follows:  similar to \cite{A}, we first reduce the satisfiability problem of $\strint$ to the corresponding satisfiability problem of the theory of Presburger Arithmetic with exponential functions (denoted by $\paexp$); then, according to the decidability of $\paexp$ \cite{Semenov84}, we therefore obtain the decidability of the satisfiability of $\strint$.

In \cite{Semenov84}, Semen\"{o}v just proved the decidability and did not provide an explicit decision procedure. So, in  \cite{Point86}, Point presented a first decision procedure for the satisfiability of $\paexp$ in a mathematical way, which is difficult to understand and therefore no implementation available based on which up to now to the best of our knowledge. So, another contribution of this paper is that we present a more understandable algorithm for $\paexp$'s satisfiability in a computational way and its prototypical implementation. Although the general framework of our algorithm is quite similar to Point's,  it contains the following improvements and corrections:

\wuhao{to do}

### 2021-05-14 FMCAD2021/4_dec_proc.tex
sec6 half done

\iffalse
\subsection{Complexity Analysis and Case Study}
In this section, we give a complexity analysis of our decision procedure, and provide an example to demonstrate the procedure.
%\subsection{Complexity Analysis}

@@

%\subsection{Example}

We apply our algorithm to a simple example to 
illustrate its function.
Given $\Sigma = \{0,1\}$
Suppose $\mathcal{A}$ is a finite state automaton
that recognizes language $L = \{ w \mid \#w('1') \text{ is odd}\}$.
Consider the following string constraint
$$y\in L(\mathcal{A})\wedge 4\le  \parseInt(y) < 16\,.$$

We need an abstraction parameter $alpha$ to restrict interpretations
for $x$, for simplicity, set $\alpha\Def \langle 2,1 \rangle$.
The generic $\langle 2,1 \rangle$-flat language is the union of 
$\langle 2,1 \rangle$-flat languages $(11)^*,(10)^*,(01)^*,(00)^*,1^*,0^*$.
Here we divide the problem into 6 sub-cases by restricting $y$
to one of the $\langle 2,1 \rangle$-flat language.
Another approach is to directly restrict $y$ to 
the generic $\langle 2,1 \rangle$-flat language,
at the cost of more introduced variables.
Since the complexity is related to the number of variables,
we wish it to be as little as possible.

When $y$ is in language $1^*$, assume $y = 1^x$ where $x$
is a fresh integer variable.
Note that the Parikh image of $L(\mathcal{A})$ is $2|( x+1)$, 
so the constraint is converted to a PA formula $\phi$
$$\phi\Def \exists x. 2|(x+1) \wedge 5 \le 2^{x} < 17\,.$$

Observe that when $x=3$ the formula holds,
so we can easily deduce that $\phi$ is true.  
By applying our algorithm,
after \textbf{Normalization},
$\phi$ is transformed into the form

$$\exists x_0. 2|(x+1)\wedge  2^{x}\le 16 \wedge -2^{x}\le -5\,.$$

Since there is only one variable, 
\textbf{Enumerate-Orders} can be omitted.
We send $2^x\le 16$ and $-2^x \le -5$ to \textbf{Elim-exp} and obtain
$$2^x\le 16 \equiv x \le 4$$
$$-2^x\le -5 \equiv 3\le x\,.$$ 
Then we apply \textbf{Elim-linear} on formula $\exists x_0. 2|(x+1) \wedge x\le 4 \wedge 3\le x$ and get
$$\bigvee_{j=0}^1(2| j+1 \wedge \textit{false} \wedge \textit{true})\vee \bigvee_{j=0}^1 (2|4-j+i\wedge 4-j\le 4 \wedge 3\le 4-j) \,.$$
which evaluates \textit{true} and we can deduce $j=1\implies x=3$, so $y$ is string $111$.

Other sub-cases are similar. However, when the abstraction parameter becomes larger,
say, $\alpha =\langle 2,2 \rangle$,
the complexity will grow rapidly,
because \textbf{Enumerate-Orders} can not always be omitted and formulas in \textbf{Elim-exp} can not be simplified easily.
\fi

>>>>

\subsection{Complexity Analysis}
\wuhao{To be checked}
In this section, we give a complexity analysis of our decision procedure, and provide an example to demonstrate the procedure.
%\subsection{Complexity Analysis}

@@

### 2021-05-14 FMCAD2021/6_impl_exp.tex
implement ongoing
sec6 half done

We have implemented a prototype of our algorithm in Wolfram Mathematica, called  $\paexp$-Solver. Any postive Integer is allowed to be the base of exponential functions in our algorithm, but $\parseInt$ in Z3 and CVC4 only takes a numeric string as input. For this reason, we fix the exponential function to be $10^x$ in our experiments.

@@

\paragraph{string hash function}
 flat pattern and arbitrarys pattern

All experiments were run on an Intel Core i5 1.4GHz with 8GB RAM.

In our benchmark, the coefficients in inequalities are generated randomly within a range. As a consequence, some inequalities may not actually work, for example, $-10^{x_1} - x_1 \le 1$. So we always set the number of exponential  inequalities is one more than number of exponential variables, and the same for the linear comterpart. Since exponential terms with large coefficients are more likely to dominate the summation, coefficients of exponential terms are in the range $[-10,10]$ while other coefficients and constarints are generated randomly in $[-10^5,10^5]$.

For each

>>>>

\subsection{Implementation}
We have implemented a prototype of our algorithm in Wolfram Mathematica, called  $\paexp$-Solver. Any postive Integer is allowed to be the base of exponential functions in our algorithm, but $\parseInt$ in Z3 and CVC4 only takes a numeric string as input. For this reason, we fix the exponential function to be $10^x$ in our experiments.

@@

\paragraph{string hash function}

In the sequel, we consider string hash functions of the form  
\begin{equation}
    \text{hash}(x)= \parseInt(x) \mod m_1  \notag
\end{equation}
where $m_1$ is usually a large (prime) number. We would like to find if there exists strings that match desired patterns and are hashed into certain integer values. In our benchmark, we focus string constraints of the form
\begin{equation}
    \left( \parseInt(x) \mod m_1\right) \mod m_2 = 0   \notag
\end{equation}
where $x$ either has a prefix ``12345'' or a suffix ``6789'' or both, for example $x=``12345"y``6789"$. We choose 2 patterns for $y$ to match:

The first pattern is a FL with 2 cycles, $y \in (a)^+(b)^+$. Here we assume each cycle repeat at least once so that it will not degrade into FL with 1 cycle. The constraints is encoded to a $\paexp$ formula where two exponential variables denote the number of cycle $a$ and $b$ respectively.\wuhao{Need example?}

The Second pattern is arbitrary string, that is $y\in \Sigma_{num}^*$. We use languages of the form $(a^1_1)^* (a^2_1...a^2_l)$ to under-approximate arbitrary strings, the first part is a single digit and the second part is a string length $l$, which can be seen as a FL with 2 cycles and force the second cycle do not repeat. We first set $l=1$ and enumerate $a^1_1$ from $1$ to $9$, for each case we obtain an arithmetic formula with $1$ exponential variables. If no satisfying model is found, we increase $l$ by 1 and repeat this procedure until it finds a model or timeout. This strategy is similar to \cite{Abdulla2020} with the difference that they force $a^1_1=0$ so that no exponential terms will occur.

\subsection{Experiment Results}

All experiments are run on a Intel i5 1.4GHz with 8GB RAM with TIMEOUT set to be 60 seconds.

\wuhao{where to place tables?}

From Table \ref{table:arithmetic}, we can see that for all cases except bounded 4-5-5-6 case, $\paexp$-Solver answers more \textit{sat} and \textit{unsat} than Z3 and CVC4. Our solver significantly outperforms when equalities are included. The average time needed to assert \textit{sat} or \textit{unsat} increase when we add more exponential variables.\wuhao{more details needed}

For string hash function benchmark, we did compared our solver with Z3, CVC4 and Trau, but the experiments show that Trau would incorrectly answer \textit{unsat} in most cases even though a model can be found. So in \ref{table:string}, we still show the results of the other 3 solvers. As we can see, for flat patterns $(a)^*(b)^*$, Z3 and CVC4 would run out of time on most cases while $\paexp$-solver can give an answer for almost all trails. For arbitrary strings $(\Sigma_{num})^*$, CVC4 and Z3 seem to be biased against trails with surfix or prefix. By applying the under-approximation technique, our tool is able to find a match on all trails. \wuhao{more details needed}

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
\begin{table*}[]
    \label{table:arithmetic}
    \caption{Results of Z3, CVC4 and $\paexp$-Solver on arithmetic benchmark}
    \begin{tabular}{cc|cc|cc|cc}
    \multicolumn{1}{l}{} &  & \multicolumn{2}{c|}{Z3} & \multicolumn{2}{c|}{CVC4} & \multicolumn{2}{c}{MySolver} \\
    \multicolumn{1}{l}{} &  & number & avg & number & avg & number & avg \\ \hline
    \multirow{3}{*}{\begin{tabular}[c]{@{}c@{}}unbound\\ 2-3-3-4\end{tabular}} & SAT & 31 & \textless{}0.1 & 26 & 1.2 & 35 & \textless{}0.1 \\
     & UNSAT & 37 & \textless{}0.1 & 38 & \textless{}0.1 & 65 & 0.3 \\
     & TIMEOUT & 32 &  & 36 &  & 0 &  \\ \hline
    \multirow{3}{*}{\begin{tabular}[c]{@{}c@{}}unbound\\ 3-4-4-5\end{tabular}} & SAT & 27 & \textless{}0.1 & 16 & \textless{}0.1 & 34 & 0.4 \\
     & UNSAT & 22 & \textless{}0.1 & 24 & \textless{}0.1 & 66 & 2.8 \\
     & TIMEOUT & 51 &  & 60 &  & 0 &  \\ \hline
    \multirow{3}{*}{\begin{tabular}[c]{@{}c@{}}unbound\\ 3-4-4-5 \\ with equality\end{tabular}} & SAT & 4 & 19 & 0 & \textless{}0.1 & 16 & 3.7 \\
     & UNSAT & 30 & \textless{}0.1 & 31 & 0.2 & 44 & \textless{}0.1 \\
     & TIMEOUT & 66 &  & 69 &  & 28 & 19.6 \\ \hline
    \multirow{3}{*}{\begin{tabular}[c]{@{}c@{}}unbound\\ 4-5-5-6\end{tabular}} & SAT & 20 & 0.2 & 15 & 8.2 & 33 & 5.2 \\
     & UNSAT & 22 & \textless{}0.1 & 23 & 0.2 & 50 & 4.8 \\
     & TIMEOUT & 58 &  & 62 &  & 17 &  \\ \hline\hline
    \multirow{4}{*}{\begin{tabular}[c]{@{}c@{}}bound\\ 2-3-3-4\end{tabular}} & SAT & 31 & \textless{}0.1 & 14 & \textless{}0.1 & 37 & 0,6 \\
     & UNSAT & 31 & \textless{}0.1 & 30 & \textless{}0.1 & 46 & \textless{}0.1 \\
     & B-UNSAT &  &  &  &  & 17 & 9.3 \\
     & TIMEOUT & 38 &  & 56 &  & 0 &  \\ \hline
    \multirow{4}{*}{\begin{tabular}[c]{@{}c@{}}bound\\ 3-4-4-5\end{tabular}} & SAT & 31 & 0.8 & 13 & \textless{}0.1 & 36 & 3.2 \\
     & UNSAT & 20 & \textless{}0.1 & 20 & \textless{}0.1 & 42 & \textless{}0.1 \\
     & B-UNSAT &  &  &  &  & 3 & 54 \\
     & TIMEOUT & 49 &  & 67 &  & 19 &  \\ \hline
    \multirow{4}{*}{\begin{tabular}[c]{@{}c@{}}bound\\ 3-4-4-5 \\ with equality\end{tabular}} & SAT & 3 & 23 & 1 & 39 & 14 & 27.5 \\
     & UNSAT & 25 & \textless{}0.1 & 25 & \textless{}0.1 & 47 & \textless{}0.1 \\
     & B-UNSAT &  &  &  &  & 2 & 57.9 \\
     & TIMEOUT & 72 &  & 74 &  & 37 &  \\ \hline
    \multirow{4}{*}{\begin{tabular}[c]{@{}c@{}}bound\\ 4-5-5-6\end{tabular}} & SAT & 29 & 2 & 11 & \textless{}0.1 & 28 & 10 \\
     & UNSAT & 20 & \textless{}0.1 & 20 & \textless{}0.1 & 38 & \textless{}0.1 \\
     & B-UNSAT &  &  &  &  & 0 &  \\
     & TIMEOUT & 51 &  & 69 &  & 34 & 
    \end{tabular}
\end{table*}

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
\begin{table*}[]
    \label{table:string}
    \caption{Results of Z3, CVC4 and $\paexp$-Solver on string hash function benchmark}
    \begin{tabular}{c|c|ccc|ccc|ccc}
                                                       &     & \multicolumn{3}{c|}{Z3} & \multicolumn{3}{c|}{CVC4} & \multicolumn{3}{c}{MySolver}       \\
    PATTERN                                            & $\ell_{10}(m_1)$   & SAT  & UNSAT  & TIMEOUT & SAT   & UNSAT  & TIMEOUT  & SAT & UNSAT & TIMEOUT              \\ \hline
    \multirow{5}{*}{12345(a)*(b)*}                     & 2   & 5    &        & 20      & 23    &        & 2        & 24  & 1     & 0                    \\
                                                       & 3   & 0    &        & 25      & 4     &        & 21       & 10  & 14    & 1                    \\
                                                       & 4   & 0    &        & 25      & 1     &        & 24       & 2   & 22    & 1                    \\
                                                       & 5   & 0    &        & 25      & 1     &        & 24       & 1   & 23    & 1                    \\
                                                       & total & 5    &        & 95      & 29    &        & 71       & 37  & 60    & 3                    \\ \hline
    \multirow{5}{*}{12345(a)*(b)*6789}                 & 2   & 10   &        & 15      & 18    &        & 7        & 24  & 1     & 0                    \\
                                                       & 3   & 1    &        & 24      & 8     &        & 17       & 10  & 15    & 0                    \\
                                                       & 4   & 0    &        & 25      & 3     &        & 22       & 3   & 22    & 0                    \\
                                                       & 5   & 0    &        & 25      & 0     &        & 25       & 0   & 25    & 0                    \\
                                                       & total & 11   &        & 89      & 29    &        & 71       & 37  & 63    & 0                    \\ \hline
    \multirow{5}{*}{(a)*(b)*6789}                      & 2   & 14   &        & 11      & 23    &        & 2        & 22  & 3     & 0                    \\
                                                       & 3   & 4    &        & 21      & 6     & 1      & 18       & 15  & 10    & 0                    \\
                                                       & 4   & 0    &        & 25      & 1     &        & 24       & 3   & 22    & 0                    \\
                                                       & 5   & 0    &        & 25      & 0     &        & 25       & 1   & 24    & 0                    \\
                                                       & total & 18   &        & 82      & 30    & 1      & 69       & 41  & 59    & 0                    \\ \hline\hline
    \multirow{5}{*}{12345$(\Sigma_{num})^*$}     & 2   & 25   &        & 0       & 25    &        & 0        & 25  &       &                      \\
                                                       & 3   & 25   &        & 0       & 25    &        & 0        & 25  &       &                      \\
                                                       & 4   & 23   &        & 2       & 25    &        & 0        & 25  &       &                      \\
                                                       & 5   & 9    &        & 16      & 25    &        & 0        & 25  &       &                      \\
                                                       & total & 82   &        & 18      & 100   &        & 0        & 100 &       &                      \\ \hline
    \multirow{5}{*}{12345$(\Sigma_{num})^*$6789} & 2   & 25   &        & 0       & 15    &        & 10       & 25  &       & \multicolumn{1}{l}{} \\
                                                       & 3   & 23   &        & 2       & 1     &        & 24       & 25  &       & \multicolumn{1}{l}{} \\
                                                       & 4   & 9    &        & 16      & 1     &        & 24       & 25  &       & \multicolumn{1}{l}{} \\
                                                       & 5   & 3    &        & 22      & 0     &        & 25       & 25  &       &                      \\
                                                       & total & 60   &        & 40      & 17    &        & 83       & 100 &       &                      \\ \hline
    \multirow{5}{*}{$(\Sigma_{num})^*$6789}      & 2   & 25   &        & 0       & 19    &        & 6        & 25  &       & \multicolumn{1}{l}{} \\
                                                       & 3   & 25   &        & 0       & 4     &        & 21       & 25  &       & \multicolumn{1}{l}{} \\
                                                       & 4   & 16   &        & 9       & 4     &        & 21       & 25  &       & \multicolumn{1}{l}{} \\
                                                       & 5   & 2    &        & 23      & 0     &        & 25       & 25  &       & \multicolumn{1}{l}{} \\
                                                       & total & 68   &        & 32      & 27    &        & 73       & 100 &       &                     
    \end{tabular}
\end{table*}
